---
title: "SCRNAnalysis workflow with a larger annotated dataset"
author: "Giuseppe Bocci"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{SCRNAnalysis workflow with a larger annotated dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

required_packages <- c(
  "SCRNAnalysis", "scRNAseq", "SingleCellExperiment",
  "SummarizedExperiment", "S4Vectors", "scater", "SingleR",
  "ggplot2", "ggdendro", "ggalluvial"
)
packages_available <- vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)

# ZeiselBrainData is downloaded through ExperimentHub on first use. Keep the
# package build self-contained; set this environment variable to render all
# computations and figures locally.
run_large_example <- identical(
  Sys.getenv("SCRNANALYSIS_RUN_LARGE_VIGNETTE"), "true"
)
workflow_eval <- all(packages_available) && run_large_example
```

```{r missing-packages, echo=FALSE, results='asis'}
if (!all(packages_available)) {
  cat(
    "> Install the optional packages required for this workflow: ",
    paste(names(packages_available)[!packages_available], collapse = ", "),
    ".\n\n",
    sep = ""
  )
}
```

# Introduction

This vignette presents a larger, practical workflow based on the annotated mouse brain dataset from Zeisel et al. The dataset contains thousands of cells and several broad cell classes. It is obtained through `scRNAseq::ZeiselBrainData()`, so no large data files are included in the package source.

The cells are randomly divided into a query set and an annotated reference set. The query is filtered, normalized, clustered, and annotated with SingleR using the reference labels. This separation avoids evaluating the labels of a cell against itself.

## Note on vignette evaluation
If vignette does not produce the figures locally, set the following before building the vignette:

```{r enable-large-workflow, eval=FALSE}
Sys.setenv(SCRNANALYSIS_RUN_LARGE_VIGNETTE = "true")
devtools::build_vignettes(".")
```

# Load the annotated dataset

```{r load-data, eval=workflow_eval, message=FALSE, warning=FALSE}
library(SCRNAnalysis)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(S4Vectors)

brain <- scRNAseq::ZeiselBrainData()
labels <- SummarizedExperiment::colData(brain)$level1class

keep <- !is.na(labels)
brain <- brain[, keep]
labels <- droplevels(factor(labels[keep]))

table(labels)
dim(brain)
```

# Create independent query and reference sets

```{r split-data, eval=workflow_eval}
set.seed(2026)
query_cells <- sample(seq_len(ncol(brain)), size = floor(0.4 * ncol(brain)))

query_sce <- brain[, query_cells]
reference_sce <- brain[, -query_cells]
SummarizedExperiment::colData(reference_sce)$cell_type <- labels[-query_cells]

# SingleR expects a log-expression assay in the reference.
reference_sce <- scater::logNormCounts(reference_sce)
query <- ScDeconvExperiment(query_sce)
```

# Filter, normalize, and identify variable genes

```{r preprocess-query, eval=workflow_eval}
query <- filterLowExpressionGenes(
  query,
  assay_name = "counts",
  min_count = 1,
  lower_percentile = 0.01
)
query <- filterCellsByDetectedGenes(
  query,
  assay_name = "counts",
  min_count = 1,
  lower_percentile = 0.01,
  upper_percentile = 0.99
)
query <- normalizeToTotal(query, target_total = 1e4)
query <- logTransform(query, assay_name = "normcounts")

n_hvg <- min(2000, nrow(query))
query <- identifyHVGs(query, assay_name = "logcounts", n_top = n_hvg)
```

# Cluster and annotate the query cells

```{r cluster-and-annotate, eval=workflow_eval, message=FALSE, warning=FALSE}
n_pcs <- min(30, ncol(query) - 1, sum(SummarizedExperiment::rowData(query)$is_hvg) - 1)
n_pcs <- max(2, n_pcs)

query <- performClustering(
  query,
  assay_name = "logcounts",
  algorithm = "kmeans",
  use_hvg = TRUE,
  n_hvg = n_hvg,
  n_pcs = n_pcs,
  n_clusters = 10,
  cluster_column = "cluster"
)

query <- performDeconvolution(
  query,
  reference = reference_sce,
  flavour = "singler",
  reference_label_column = "cell_type",
  assay_name = "logcounts",
  ref_assay_name = "logcounts",
  label_column = "predicted_cell_type"
)
```

# Compare clusters and annotations

```{r compare, eval=workflow_eval}
query <- compareClusterCellTypes(
  query,
  cluster_column = "cluster",
  cell_type_column = "predicted_cell_type",
  metadata_name = "large_dataset_comparison"
)
```

```{r heatmap, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeHeatmap(
  query,
  metadata_name = "large_dataset_comparison",
  value = "row_proportions"
)
```

```{r dendrogram, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeDendrogram(
  query,
  metadata_name = "large_dataset_comparison",
  value = "row_proportions",
  margin = "clusters"
)
```

```{r alluvial, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeAlluvial(
  query,
  metadata_name = "large_dataset_comparison"
)
```

# Reproducibility information

```{r history, eval=workflow_eval}
query@history
str(query@params, max.level = 2)
```
