---
title: "SCRNAnalysis workflow with Seurat data"
author: "Giuseppe Bocci"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{SCRNAnalysis workflow with Seurat data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

required_packages <- c(
  "SCRNAnalysis",
  "Seurat",
  "SeuratObject",
  "SingleCellExperiment",
  "SummarizedExperiment",
  "S4Vectors",
  "scater",
  "ggplot2",
  "ggdendro",
  "ggalluvial"
)

packages_available <- vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
workflow_eval <- all(packages_available)
```

```{r missing-packages, echo=FALSE, results='asis'}
if (!workflow_eval) {
  missing_packages <- names(packages_available)[!packages_available]
  cat(
    "> Some chunks are not evaluated because the following packages are missing: ",
    paste(missing_packages, collapse = ", "),
    ".\n\n",
    sep = ""
  )
}
```

# Introduction

This vignette shows a complete `SCRNAnalysis` workflow using `pbmc_small`, a small dataset included in `Seurat`. The dataset is used as a lightweight example to demonstrate the main analysis steps without requiring external downloads.

The workflow includes:

1. loading data from a `Seurat` object;
2. converting the data to `SingleCellExperiment`;
3. creating a `ScDeconvExperiment` object;
4. filtering genes and cells;
5. normalizing counts and applying a log-transform;
6. selecting highly variable genes;
7. clustering cells;
8. annotating cells using Seurat label transfer;
9. comparing clusters with predicted cell-type labels;
10. visualizing the relationship between clusters and annotations.

`pbmc_small` is suitable for a vignette because it is very small. The labels stored in its metadata are used here only to demonstrate the workflow.

# Load packages

```{r packages, eval=workflow_eval, message=FALSE, warning=FALSE}
library(SCRNAnalysis)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(S4Vectors)
```

# Load Seurat data

`pbmc_small` is a `Seurat` object included in the `Seurat` package. In this vignette, it is used both as the source of query counts and as an annotated reference for label transfer.

```{r load-seurat-data, eval=workflow_eval, message=FALSE, warning=FALSE}
pbmc_small <- SeuratObject::pbmc_small
reference <- Seurat::UpdateSeuratObject(pbmc_small)
reference
```

The object metadata contains columns that can be used as known labels. Here, `letter.idents` is used as the reference label column.

```{r inspect-reference-labels, eval=workflow_eval}
head(reference[[]])
table(reference$letter.idents)
```

To make the reference usable for Seurat label transfer, we apply minimal preprocessing.

```{r preprocess-reference, eval=workflow_eval, message=FALSE, warning=FALSE}
reference <- Seurat::NormalizeData(reference, verbose = FALSE)
reference <- Seurat::FindVariableFeatures(reference, verbose = FALSE)
reference <- Seurat::ScaleData(reference, verbose = FALSE)
reference <- Seurat::RunPCA(reference, verbose = FALSE)
```

# Create a `ScDeconvExperiment` object

The `ScDeconvExperiment()` constructor defined in the package expects a `SingleCellExperiment` object containing a `counts` assay. For this reason, we first extract the counts from the `Seurat` object, then create a `SingleCellExperiment`, and finally convert it to the package-specific class.

The helper function `get_seurat_assay_data()` keeps the code compatible with both Seurat v5, which uses `layer`, and earlier versions, which use `slot`.

```{r create-object, eval=workflow_eval, message=FALSE, warning=FALSE}
get_seurat_assay_data <- function(x, assay = SeuratObject::DefaultAssay(x), layer = "counts") {
  tryCatch(
    SeuratObject::GetAssayData(x, assay = assay, layer = layer),
    error = function(e) SeuratObject::GetAssayData(x, assay = assay, slot = layer)
  )
}

counts <- get_seurat_assay_data(reference, assay = "RNA", layer = "counts")
query_sce <- SingleCellExperiment::SingleCellExperiment(assays = list(counts = counts))
query <- ScDeconvExperiment(query_sce)
query

dim(SummarizedExperiment::assay(query, "counts"))
```

# Filter genes and cells

The first quality-control step removes genes expressed in too few cells. The second step removes cells with an abnormal number of detected genes, using percentile-based thresholds.

```{r quality-control, eval=workflow_eval}
query <- filterLowExpressionGenes(
  query,
  assay_name = "counts",
  min_count = 1,
  lower_percentile = 0.01
)
query <- filterCellsByDetectedGenes(
  query,
  assay_name = "counts",
  min_count = 1,
  lower_percentile = 0.01,
  upper_percentile = 0.99
)
query@params$filterLowExpressionGenes
query@params$filterCellsByDetectedGenes
dim(SummarizedExperiment::assay(query, "counts"))
```

The metrics computed during filtering are stored in `rowData()` and `colData()`.

```{r qc-metadata, eval=workflow_eval}
head(SummarizedExperiment::rowData(query))
head(SummarizedExperiment::colData(query))
```

# Normalize counts and apply a log-transform

After filtering, counts are normalized to a total of 10,000 counts per cell and stored in `normcounts`. A log-transform is then applied and stored in `logcounts`.

```{r normalization, eval=workflow_eval}
query <- normalizeToTotal(query, target_total = 1e4)
query <- logTransform(query, assay_name = "normcounts", pseudo_count = 1, base = 2)

SummarizedExperiment::assayNames(query)
head(SummarizedExperiment::colData(query)$library_size)
```

# Select highly variable genes

Highly variable genes are selected from the `logcounts` assay. Since this example dataset is small, we select at most 100 genes.

```{r hvg, eval=workflow_eval}
n_hvg <- min(100, nrow(query))
query <- identifyHVGs(query, assay_name = "logcounts", n_top = n_hvg, method = "dispersion")

table(SummarizedExperiment::rowData(query)$is_hvg)
head(SummarizedExperiment::rowData(query)[order(SummarizedExperiment::rowData(query)$hvg_rank), ])
```

# Cluster cells

`performClustering()` first computes a PCA using `scater::runPCA()` and then applies the selected clustering algorithm. Here, k-means clustering is used with seven clusters so that the comparison plots show a more detailed relationship with the reference annotations.

```{r clustering, eval=workflow_eval, message=FALSE, warning=FALSE}
n_pcs <- min(5, ncol(query) - 1, sum(SummarizedExperiment::rowData(query)$is_hvg) - 1)
n_pcs <- max(2, n_pcs)

query <- performClustering(
  object = query,
  assay_name = "logcounts",
  algorithm = "kmeans",
  use_hvg = TRUE,
  n_hvg = n_hvg,
  n_pcs = n_pcs,
  n_clusters = 7,
  cluster_column = "cluster",
  pca_name = "PCA"
)

table(SummarizedExperiment::colData(query)$cluster)
```

# Annotate cells with Seurat

Cell labels are transferred from the Seurat reference to the query object. `performDeconvolution()` stores the predicted labels in `colData(query)$seurat_label` and the maximum prediction score in `colData(query)$seurat_label_score`.

```{r seurat-annotation, eval=workflow_eval, message=FALSE, warning=FALSE}
dims <- seq_len(min(10, ncol(SeuratObject::Embeddings(reference, "pca"))))

query <- performDeconvolution(
  object = query,
  reference = reference,
  flavour = "seurat",
  reference_label_column = "letter.idents",
  label_column = "seurat_label",
  dims = dims
)

table(SummarizedExperiment::colData(query)$seurat_label)
summary(SummarizedExperiment::colData(query)$seurat_label_score)
```

# Compare clusters and cell-type labels

`compareClusterCellTypes()` builds a contingency table between the computed clusters and the predicted cell-type labels. The result is stored in `metadata(query)$cluster_cell_type_comparison`.

```{r cluster-cell-type-comparison, eval=workflow_eval}
query <- compareClusterCellTypes(
  object = query,
  cluster_column = "cluster",
  cell_type_column = "seurat_label",
  metadata_name = "cluster_cell_type_comparison"
)

comparison <- S4Vectors::metadata(query)$cluster_cell_type_comparison
head(comparison$long_table)
```

# Visualize the results

The package provides three plotting functions to explore the relationship between clusters and annotations.

```{r heatmap, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeHeatmap(query, metadata_name = "cluster_cell_type_comparison", value = "row_proportions")
```

```{r dendrogram, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeDendrogram(query, metadata_name = "cluster_cell_type_comparison", value = "row_proportions", margin = "clusters")
```

```{r alluvial, eval=workflow_eval, message=FALSE, warning=FALSE}
plotClusterCellTypeAlluvial(query, metadata_name = "cluster_cell_type_comparison")
```

# Workflow history and parameters

The `ScDeconvExperiment` class keeps track of the parameters used by the workflow and of the analysis steps that were executed. This helps make the workflow more reproducible.

```{r workflow-history, eval=workflow_eval}
query@history
str(query@params, max.level = 2)
```

# SingleR alternative

The workflow above uses `flavour = "seurat"`. The same function can also be used with `flavour = "singler"`, provided that the reference is a `SummarizedExperiment` or `SingleCellExperiment` object with an annotation column in `colData()`.

This chunk is not evaluated by default because the main vignette uses Seurat, but it shows the equivalent call for SingleR.

```{r singler-alternative, eval=FALSE}
reference_sce <- Seurat::as.SingleCellExperiment(reference)

if (!"logcounts" %in% SummarizedExperiment::assayNames(reference_sce)) {
  ref_logcounts <- get_seurat_assay_data(reference, assay = "RNA", layer = "data")
  SummarizedExperiment::assay(reference_sce, "logcounts") <- ref_logcounts
}

SummarizedExperiment::colData(reference_sce)$cell_type <- reference$letter.idents

query <- performDeconvolution(
  object = query,
  reference = reference_sce,
  flavour = "singler",
  reference_label_column = "cell_type",
  assay_name = "logcounts",
  ref_assay_name = "logcounts",
  label_column = "singler_label"
)

table(SummarizedExperiment::colData(query)$singler_label)
```
